Spatio-angular fluorescence microscopy I. Basic theory

### 2018-10-20 papers/basic-theory/paper/paper.tex
Broad updates to paper.

% Talon's custom commands
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}[1]{\mathbf{\mathbf{r}}_o}
\providecommand{\so}[1]{\mathbf{\hat{s}}_o}
\providecommand{\rb}[1]{\mathbf{r}_b}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}[1]{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\DeclareFontFamily{OT1}{pzc}{}
\DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}

@@

\begin{abstract*}
  We introduce the basic elements of a complete spatio-angular theory of
  fluorescence microscopes. We start by analyzing a single-view fluorescence
  microscope imaging an ensemble of in-focus dipole absorber/emitters using
  electromagnetic optics and the paraxial approximation. Next, we define the
  spatio-angular transfer function and show that fluorescence microscopes have
  an angular band limit. Finally, we discuss the implications of our analysis on
  quantitative fluorescence microscopy studies.
\end{abstract*}

%%%%%%%%%%%%%%%%%%%%%%%%%%  body  %%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Fluorescence microscopes are ubiquitous tools in biology [XXX], materials
science [XXX], and metrology [XXX] for measuring the concentration of
fluorophores throughout a sample. While an unprocessed fluorescence micrograph
reports the approximate concentration of fluorophores throughout a sample,
diffraction-limited microscopes image a spatially blurred version of the true
fluorophore distribution. Therefore, microscopists who are interested in
accurate measurements will perform a computational restoration to recover some
of the information lost during the imaging process.

Restoration techniques attempt to recover the true distribution of fluorophores
using the measured data and a model of the imaging process. The model of the
imaging process can be obtained theoretically (by mathematically modeling the
instrument under idealized conditions), experimentally (by measuring the
microscope's response to a known input), or by a combination of theory and
experiment (by measuring parameters of an instrument model). In all cases, the
accuracy of the restored image is limited by the accuracy of the imaging model.

Most theoretical fluorescence microscopy imaging models make simplifying
approximations. A common (often unstated) pair of approximations is to (1) treat
fluorophores as monopole absorber/emitters and (2) treat light propagation under
the scalar approximation. In this work we will analyze fluorescence microscopes
without these approximations and find the conditions under which these
approximations are valid.

Using an experimentally determined model is an effective way to find an accurate
model without understanding all of the physical details of the imaging system,
but these approaches require a known input that can completely characterize the
imaging system. Sub-diffraction beads are commonly used as a known input for
measuring the response of a microscope, but we argue that sub-diffraction beads
are not enough to completely characterize the response of a microscope to all
fluorescent samples. Beads can only characterize the response of the microscope
to an angularly uniform distribution of fluorophores, and we will show that all
microscopes have an orientation-dependent response. A complete characterization
requires a test sample with at least three known orientation distributions.

The single-molecule localization microscopy (SMLM) community has pioneered the
use of rigorous electromagnetic models of single fluorescence microscopes
\cite{backer2014, lieb2004}. When only one fluorophore is emitting at a time,
the measured intensity pattern is strongly dependent on the orientation of the
emitting molecule. Backlund and Lew \cite{backlund2014} have shown that ignoring
the orientation of fluorophores can lead to biased estimates of their position,
so the most accurate SMLM experiments must jointly estimate the position and
orientation of the molecule. These joint estimates usually use a set of
precomputed intensity patterns and a maximum-likelihood algorithm to estimate
the orientation and position of each molecule. These methods have led to some of
the most accurate position estimates of single molecules.

Meanwhile, most fluorescence microscopists image ensembles of fluorophores.
Although many groups have used polarizers to probe the orientation of ensembles
of fluorophores \cite{mehta2016} [Forkey], to our knowledge no
work has been done on the response of fluorescence microscopes to ensembles of
oriented fluorophores. Furthermore, current restoration techniques in polarized
light microscopy are lacking because they rely on simple pixel-wise arithmetic
and do not model the propagation of light through the microscope.

In this work we model ensembles of in-focus dipole absorber/emitters using
electromagnetic optics and the paraxial approximation. This marks a first step
towards a complete three-dimensional non-paraxial theory of fluorescence
microscopes. In section 2 we model the excitation and detection processes of
dipoles. We define the most important quantity in this work, the spatio-angular
transfer function, and we use it to show that fluorescence microscopes have an
angular band limit. In section 3 we perform simulations and compare our model to
traditional scalar models. In section 4 we discuss the implications of our work
for fluorescence microscopy and lay out a path towards a complete spatio-angular
theory of fluorescence microscopy.

\section{Theory}
\subsection{Spherical harmonics}\label{sec:sph}
\begin{align}
  y_l^m(\vartheta, \varphi) =
  \begin{cases}
    \sqrt{2}K_l^m\cos(m\varphi)P_l^m(\cos\vartheta), & m > 0\\
    K_l^0P_l^0(\cos\vartheta), & m = 0\\
    \sqrt{2}K_l^m\sin(-m\varphi)P_l^{-m}(\cos\vartheta), & m < 0\\
  \end{cases}
\end{align}
where
\begin{align}
  K_l^m = \sqrt{\frac{(2l+1)}{4\pi}\frac{(l-|m|)!}{(l+|m|)!}},
\end{align}
and $P_l^m(x)$ are the associated Legendre polynomials. The $l=0, 1, 2$
spherical harmonics are given by
\begin{align}
  y_0^0(\vartheta, \varphi) &= \sqrt{\frac{1}{4\pi}}, \nonumber \\ 
  y_1^{-1}(\vartheta, \varphi) = \sqrt{\frac{3}{4\pi}}\sin\varphi\sin\vartheta, \hspace{2em} y_1^0(\vartheta, \varphi) &= \sqrt{\frac{3}{4\pi}}\cos\vartheta, \hspace{2em} y_1^1(\vartheta, \varphi) = \sqrt{\frac{3}{4\pi}}\cos\varphi\sin\vartheta, \nonumber \\
  y_2^{-2}(\vartheta, \varphi) = \sqrt{\frac{15}{4\pi}}\sin\varphi\cos\varphi\sin^2\vartheta, &\hspace{2em} y_2^{-1}(\vartheta, \varphi) = \sqrt{\frac{15}{4\pi}}\sin\varphi\sin\vartheta\cos\vartheta,\nonumber \\
  y_2^0(\vartheta, \varphi) &= \sqrt{\frac{5}{16\pi}}(3\cos^2\vartheta - 1), \nonumber \\ 
  y_2^{1}(\vartheta, \varphi) = \sqrt{\frac{15}{4\pi}}\cos\varphi\sin\vartheta\cos\vartheta, &\hspace{2em} y_2^{2}(\vartheta, \varphi) = \sqrt{\frac{15}{4\pi}}(\cos^2\varphi - \sin^2\varphi)\sin^2\vartheta.\label{eq:harmonics}
\end{align}

\subsection{Excitation model}
% [theta**2*cos(phi - phip)**2/2 + 0.5*sin(phip)**2 + 0.5*cos(phip)**2, 0, 0, 0, sqrt(15)*sin(2*phip)/10, -sqrt(15)*theta*sin(phip)*cos(phi - phip)/5, sqrt(5)*theta**2*cos(phi - phip)**2/5 - sqrt(5)*sin(phip)**2/10 - sqrt(5)*cos(phip)**2/10, -sqrt(15)*theta*cos(phip)*cos(phi - phip)/5, -sqrt(15)*sin(phip)**2/10 + sqrt(15)*cos(phip)**2/10]
In this section we will calculate the excitation kernel for epi-illumination
microscopes---the relative probability of exciting a molecule with a known
orientation. Our approach is similar to previous work \cite{fourkas2001,  chandler2017},
but here we consider unpolarized illumination and restrict
ourselves to the paraxial approximation.

We consider a single molecule with a fixed absorption dipole moment $\so{}$ that
we express in spherical coordinates [see Fig. 1(a)] as 
\begin{align}
  \so{} &= \cos\varphi\sin\vartheta\mh{x} + \sin\varphi\sin\vartheta\mh{y} + \cos\vartheta\mh{z}. \label{eq:spherical}
\end{align}
We can rewrite Eq. \ref{eq:spherical} in terms of spherical harmonic functions
(see Section \ref{sec:sph}) as
\begin{align}
  \so{} &=\sqrt{\frac{3}{4\pi}}\left[y_1^1(\vartheta, \varphi)\,\mh{x} + y_1^{-1}(\vartheta, \varphi)\,\mh{y} + y_1^0(\vartheta, \varphi)\, \mh{z}\right]. \label{eq:sphharm}
\end{align}
We place the molecule in the focal plane of an aplanatic and
polarization-preserving objective lens with its optical axis aligned with the
$\mh{z}$ axis. Next, we place a spatially incoherent, spatially uniform,
unpolarized thermal light source (or its image) in the back aperture of the
objective lens to illuminate the focal plane. In this geometry each point in the
back focal plane generates a plane wave that illuminates the sample.

To model the unpolarized light source we will use polarized ray tracing [cite
Torok] to find the response for a single polarized ray, then integrate over the
rays and polarizations to find the complete response. First, we model the
electric field at every point on the back focal plane as
   \mb{E}_{\text{bfp}}(\phi_{\text{pol}}) &\propto \cos\phi_{\text{pol}}\mh{x} + \sin\phi_{\text{pol}}\mh{y}, \label{eq:bfp0}
where $\phi_{\text{pol}}$ is the polarization orientation and \{$\mh{x}$,
$\mh{y}$\} are transverse Cartesian basis vectors. Notice that Eq. \ref{eq:bfp0}
describes the incoherent electric fields at every point in the back focal plane,
it does not describe a coherent plane wave. Immediately after the lens the
electric field is
  \mb{E}_{\text{ff}}(\theta, \phi, \phi_{\text{pol}}) \propto \left\{[\mb{E}_\text{bfp}(\phi_{\text{pol}})\cdot\bsh{\phi}]\,\bsh{\phi} + [\mb{E}_\text{bfp}(\phi_{\text{pol}})\cdot\bsh{\rho}]\,\bsh{\theta}\right\}\sqrt{\cos\theta}, \label{eq:bfp2ff}
where \{$\theta$, $\phi$\} are spherical coordinates, \{$\bsh{\rho}$,
$\bsh{\phi}$\} are cylindrical basis vectors, and \{$\bsh{\theta}$,
$\bsh{\phi}$\} are spherical basis vectors defined in Figure 1(a) and related by
  \bsh{\rho} &= \cos\phi\mh{x} + \sin\phi\mh{y},\label{eq:rho}\\
  \bsh{\phi} &= -\sin\phi\mh{x} + \cos\phi\mh{y},\label{eq:phi}\\
  \bsh{\theta} &= \cos\phi\cos\theta\mh{x} + \sin\phi\cos\theta\mh{y} - \sin\theta\mh{z}.\label{eq:theta}               
In Eq. \ref{eq:bfp2ff} the
$[\mb{E}_\text{bfp}(\phi_{\text{pol}})\cdot\bsh{\phi}]\,\bsh{\phi}$ term models
s-polarized fields, the
$[\mb{E}_\text{bfp}(\phi_{\text{pol}})\cdot\bsh{\rho}]\,\bsh{\theta}$ term
models the rotation of p-polarized fields, and the $\sqrt{\cos\theta}$ term
models the apodization of an aplanatic lens. After plugging Eq. \ref{eq:bfp0}
and Eqs. \ref{eq:rho}--\ref{eq:theta} into Eq. \ref{eq:bfp2ff} and applying the
paraxial approximation, the electric field immediately after the lens is given
by
  \mb{E}_{\text{ff}}^{(p)}(\theta, \phi, \phi_{\text{pol}}) \propto \cos\phi_{\text{pol}}\mh{x} + \sin\phi_{\text{pol}}\mh{y} - \theta\cos(\phi - \phi_{\text{pol}})\mh{z}.\label{eq:eff}
To calculate the probability that a dipole oriented along $\so{}$ is excited by
a ray with electric field $\mb{E}$ we need to evaluate $|\mb{E}\cdot\so{}|^2$.
Evaluating this expression for each ray would be cumbersome if we used Eqs.
\ref{eq:spherical} and \ref{eq:eff} because we would need to simplify a large
number of trigonometric functions. Instead, we use Eqs. \ref{eq:sphharm} and
\ref{eq:eff} and the real Gaunt coefficients [XXXciteXXX] to find the products
of spherical harmonics. We find that
% [theta**2*cos(phi - phip)**2/2 + 0.5*sin(phip)**2 + 0.5*cos(phip)**2, 0, 0, 0, sqrt(15)*sin(2*phip)/10, -sqrt(15)*theta*sin(phip)*cos(phi - phip)/5, sqrt(5)*theta**2*cos(phi - phip)**2/5 - sqrt(5)*sin(phip)**2/10 - sqrt(5)*cos(phip)**2/10, -sqrt(15)*theta*cos(phip)*cos(phi - phip)/5, -sqrt(15)*sin(phip)**2/10 + sqrt(15)*cos(phip)**2/10]
  |\mb{E}_{\text{ff}}^{(p)}(\theta, \phi, \phi_{\text{pol}}) \cdot \so{}|^2 \propto\, &\left[1 + \theta^2\cos^2(\phi - \phi_{\text{pol}})\right]y_0^0(\so{}) + \frac{1}{\sqrt{5}}\left[ - 1 + 2\theta^2\cos^2(\phi - \phi_{\text{pol}})\right]y_2^{0}(\so{}) +\nonumber \\
                                   &\frac{\sqrt{15}}{10}\theta\sin\phi_{\text{pol}}\cos(\phi - \phi_{\text{pol}})y_2^{-1}(\so{}) + \frac{\sqrt{15}}{10}\theta\cos\phi_{\text{pol}}\cos(\phi - \phi_{\text{pol}})y_2^{1}(\so{}) -\nonumber \\  
                                   &\frac{\sqrt{15}}{20}\sin(2\phi_{\text{pol}})y_2^{-2}(\so{}) - \frac{\sqrt{15}}{20}\cos(2\phi_{\text{pol}})y_2^{2}(\so{}).\label{eq:integrand}
To calculate the complete excitation model we need to integrate Eq.
\ref{eq:integrand} over all polarization orientations and rays
  h_{\text{exc}}(\so{}) \propto \int_0^{\alpha}\theta d\theta\int_0^{2\pi}d\phi\int_0^{2\pi}d\phi_{\text{pol}}\, |\mb{E}_{\text{ff}}^{(p)} \cdot \so{}|^2. \label{eq:integral}
After plugging Eq. \ref{eq:integrand} into Eq. \ref{eq:integral} and evaluating
the integrals, all but two terms disappear and the orientation-dependent
excitation model is given by
  h_{\text{exc}}(\so{}) &\propto \left[1 + \frac{\alpha^2}{4}\right]y_0^0(\so{}) + \frac{1}{\sqrt{5}}\left[-1 + \frac{\alpha^2}{2}\right]y_2^0(\so{}). \label{eq:excitation}
We note that the excitation model depends on the orientation of the molecule,
but it does not depend on the position of the molecule in the focal plane. This
is a direct consequence of our illumination geometry---we have placed incoherent
light sources in the back focal plane of the objective so each point in the
focal plane within the field of view is illuminated equally.

\subsection{Detection model}
In this section we will calculate the detection kernel of an epi-detection
microscope---the irradiance on the detector due to single molecule at a known
position with known orientation. Our approach is similar to \cite{backer2014,
  nov2006, agrawal2012}, but here we restrict ourselves to the paraxial
approximation.

>>>>

% Talon's custom libraries
\usepackage{empheq}
% \usepackage{bm}

% Talon's custom commands
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\mathbf{\mathbf{r}}_o}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rb}{\mathbf{r}_b}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\bs{\nu}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
\DeclareFontFamily{OT1}{pzc}{}
\DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\newcommand{\eqname}[1]{\tag*{#1}}
\newcommand*\widefbox[1]{\fbox{\hspace{1em}#1\hspace{1em}}}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling a microscope imaging an ensemble of in-focus
  fluorescent dipoles as a linear Hilbert-space operator with domain
  $\mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ and range $\mbb{L}_2(\mbb{R}^2)$, and we
  express the operator in terms of four different sets of object-space basis
  functions. We show that the operator takes a particularly convenient form when
  expressed in a basis of spatial and spherical harmonics---a form we call the
  spatio-angular transfer function (SATF). We demonstrate our formalism by
  analyzing a paraxial single-view fluorescence microscope without using the
  monopole or scalar approximations. We show that this microscope has an angular
  band limit, and we demonstrate the value of the transfer function approach by
  efficiently simulating the imaging process with numerical phantoms. Notably,
  we show that information about the out-of-plane orientation of ensembles of
  in-focus fluorophores is imaged in paraxial widefield fluorescence
  microscopes. We discuss the implications of our analysis for all quantitative
  fluorescence microscopy studies and lay out a path towards a complete theory.
\end{abstract*}

\section{Introduction \tb{needs work}}

Fluorescence microscopes are widely used in the biological sciences for
measuring the distribution of fluorophores throughout a sample [cite]. While an
unprocessed fluorescence micrograph reports the approximate distribution of
fluorophores throughout a sample, all microscopes are diffraction limited
[cite], so the image is a blurred version of the true fluorophore distribution.
Therefore, microscopists who are interested in optimal measurements will perform
a computational restoration to recover some of the information lost during the
imaging process.

Restoration techniques attempt to recover the true distribution of fluorophores
using the measured data and a model of the imaging process. A model of the
imaging process can be obtained theoretically (by mathematically modeling the
instrument under idealized conditions), experimentally (by measuring the
instrument's response to a known input), or by a combination of theory and
experiment (by measuring parameters of an instrument model). In all cases the
accuracy of the restored fluorophore distribution is limited by the accuracy of
the imaging model. All theoretical imaging models make simplifying
approximations that limit the accuracy of restorations, so it is important to
verify that the approximations introduce an acceptable level of error. This work
investigates the errors introduced by two common approximations in models of
fluorescence microscopes---the \textit{monopole approximation} and the
\textit{scalar approximation}.

This work lies at the intersection of three classes of fluorescence microscopy:
(1) single-molecule localization microscopy (SMLM), (2) spatial ensemble
fluorescence microscopy including widefield, confocal, and light-sheet
techniques, and (3) polarized fluorescence microscopy. We briefly review these
three classes and focus on their restoration techniques and use of the monopole
and scalar approximations.

The SMLM community has pioneered the use of rigorous electromagnetic models of
fluorescence microscopes \cite{backer2014, lieb2004} [cite Novotny]. When a
single molecule is fluorescing in the sample, the measured intensity pattern is
strongly dependent on the orientation of the emitting molecule. Backlund and Lew
\cite{backlund2014} have shown that ignoring the orientation of fluorophores can
bias position estimates, so the most accurate SMLM experiments must jointly
estimate the position and orientation of each molecule.

Meanwhile, most fluorescence microscopists image ensembles of fluorophores and
restore their images without considering the role that the monopole and scalar
approximations play in the restoration process. A typical fluorescence
microscopist is only interested in the spatial distribution of fluorophores, so
they reason that they can ignore the orientation of the emitters.

A smaller community of microscopists is interested in measuring the orientation
of ensembles of fluorophores \cite{mehta2016} [Forkey, Goldman, Moerner,
Oldenbourg]. These techniques typically use polarized illumination or polarized
detection to make multiple measurements of the same object. Current angular
restoration techniques use a model of the dipole excitation and emission
processes [Fourkas] to recover the orientation of fluorophores using pixel-wise
arithmetic, but these techniques do not perform any spatial restoration so they
do not use all of the information available to the microscopist. To our
knowledge no work has been done to model the complete spatio-angular response of
fluorescence microscopes to ensembles of oriented fluorophores.

\tb{Out of date}
In this work we model ensembles of in-focus dipole absorber/emitters using
electromagnetic optics and the paraxial approximation. In section 2 we model the
excitation and detection processes of dipoles, and we define the most important
quantity in this work---the spatio-angular transfer function (SATF). We
calculate the SATF and show that fluorescence microscopes have an angular band
limit. In section 3 we perform simulations and compare our model to traditional
scalar models. In section 4 we discuss the implications of our work for
fluorescence microscopy and lay out a path towards a complete spatio-angular
theory of fluorescence microscopy.

\subsection{Monopole emission vs. dipole emission \tb{not sure where this section should live}}
Consider a single dipole radiator at the origin with its dipole moment oriented
along a direction $\so$. The dipole emitter creates a time-harmonic electric
field at a position $\mb{r}$ far from the dipole proportional to
\begin{align}
  \mb{E}_{\text{ff}}(\mb{r}, \so) \propto \frac{\text{exp}[ik|\mb{r}|]}{|\mb{r}|}\, \mh{r}\times\so\times\mh{r}.\label{eq:dip}
\end{align}
A typical approach to deriving Equation \ref{eq:dip} is to (1) use Maxwell's
equations to derive the inhomogeneous electromagnetic wave equations then (2)
use potentials to solve the wave equations with a dipole source term, then (3)
drop the terms that decay faster than $1/|\mb{r}|$ [cite Jackson, Novotny,
Teich]. Equation~\ref{eq:dip} shows that dipole radiators emit spherical
wavefronts of polarized light with amplitude proportional to $\cos\theta$ where
$\theta$ is the angle between the field point $\mb{r}$ and the dipole axis
$\so$.

Many models of fluorophores approximate the vector field on the left hand side
of Equation \ref{eq:dip} with a scalar field---the \textit{scalar
  approximation}---and drop the orientation dependence on the right hand
side---the \textit{monopole approximation}. The approximated field is given by
\begin{align}
  U_{\text{ff}}(\mb{r}) \propto \frac{\text{exp}[ik|\mb{r}|]}{|\mb{r}|}. \label{eq:mono}
\end{align}
Models that use Equation \ref{eq:mono} as a starting point do not account for
polarization or dipole orientation effects.

\section{Theory \tb{outline is okay, subsections vary, recently moved changes of basis to the front}}
We begin our analysis with the abstract Hilbert space formalism of Barrett and
Myers \cite{barrett2004}. Our first task is to formulate the imaging process as
a mapping between two Hilbert spaces $\mc{H}: \mbb{U} \rightarrow \mbb{V}$ where
$\mbb{U}$ is a set than contains all possible objects, $\mbb{V}$ is a set that
contains all possible datasets, and $\mc{H}$ is a model of the instrument that
maps between these two spaces. We denote (possibly infinite-dimensional)
Hilbert-space vectors in $\mbb{U}$ with $\mb{f}$, Hilbert-space vectors in
$\mbb{V}$ with $\mb{g}$, and the mapping between the spaces with

\begin{align}
  \mb{g} = \mc{H}\mb{f}.
\end{align}

Once we have identified the spaces $\mbb{U}$ and $\mbb{V}$, we can start
expressing the mapping between the spaces in a specific object-space and
data-space basis. In most cases the easiest mapping to find uses a
delta-function basis---we expand object and data space into delta functions then
express the mapping as an integral transform. After finding this mapping we can
start to investigate the same mapping in different bases.

The above discussion is quite abstract, but it is a powerful point of view that
will enable us to draw new conclusions about all fluorescence microscopes. In
Section \ref{sec:monopole} we will demonstrate the formalism by examining a
familiar monopole imaging model, and we will show that the point spread
function and the optical transfer function are mappings between object and data
space expressed in two different bases. In Section \ref{sec:dipole} we will
extend the monopole imaging model to dipoles and examine the mapping in four
different bases. In Section \ref{sec:para} we will examine the mappings for a
more specific case---a paraxial epi-illumination single-view fluorescence
microscope.

\subsection{Monopole imaging in different bases \tb{50\%}}\label{sec:monopole}
We start by considering a microscope that images a field of in-focus monopoles
by recording the irradiance on a two-dimensional detector. We can represent
the object as a function that assigns a real number to each point on a plane,
so we identify object space as $\mbb{U} = \mbb{L}_2(\mbb{R}^2)$. Similarly, we
have a two-dimensional detector that measures a real number at every point on 
a plane, so data space is the same set $\mbb{V} = \mbb{L}_2(\mbb{R}^2)$. Our
first task is complete---we have identified the geometry of object and data
space.

Next, we find and name the representations of our object and data in a
specific basis. In a delta function basis the object can be represented by a
function $f(\ro)$ called the \textit{monopole density}---the number of
monopoles at the two-dimensional position $\ro$ per unit area. Similarly, in a
delta function basis the data can be represented by a function $g(\rd)$ called
the \textit{irradiance}---the power received by a surface at position $\rd$
per unit area. It may seem pedantic to emphasize that these representations
are in a delta function basis, but we highlight this point to make it clear
that the delta function basis is not special---we can choose a different basis
at will.

% We start by considering a field of in-focus monopoles that have been excited by test
% a spatially uniform beam. As the fluorophores relax, they emit scalar waves in
% the form of Equation \ref{eq:mono}. The waves propagate through the microscope
% and we measure the irradiance on a two dimensional detector.

A reasonable starting point is to assume that the relationship between the
object and the data is \textit{linear}---this is usually true in fluorescence
microscopes since fluorophores emit incoherently so a scaled sum of fluorophores
will result in a scaled sum of the irradiance patterns created by the individual
fluorophores. If the mapping is linear we can write the irradiance as the
integral over a field of monopoles weighted by the detector response due to
single monopoles
\begin{align}
g(\rd) = \int_{\mbb{R}^2}d\ro\, h(\rd,\ro)f(\ro), \label{eq:fwdmono}
\end{align}
where $h(\rd{}, \ro{})$ is the kernel of the integral transform.

Most fluorescence microscopes are (at least approximately) shift-invariant,
which means that shifting the object will result in a corresponding shift of the
data. The mapping between an object and the data in a linear shift-invariant
fluorescence microscope can be written as
\begin{align}
  g(\rd) = \int_{\mbb{R}^2}d\ro\, h(\rd - \ro)f(\ro).  \label{eq:lsi}
\end{align}
When the imaging system is shift invariant the kernel $h(\rd - \ro)$ is usually
called the \textit{point spread function}.

The mapping between the object and the data in linear shift-invariant
microscopes takes a particularly simple form in a spatial harmonic basis. If we
apply the Fourier convolution theorem to Eq. \ref{eq:lsi} we find that
\begin{align}
  G(\bs{\nu}) = H(\bs{\nu})F(\bs{\nu}),\label{eq:freq}
\end{align}
where the \textit{irradiance spectrum} is given by
\begin{align}
  G(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
\end{align}
the \textit{optical transfer function} is given by
\begin{align}
  H(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
\end{align}
and the \textit{monopole spectrum} is given by
\begin{align}
    F(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}).
\end{align}
Notice that Eqs. \ref{eq:lsi} and \ref{eq:freq} are expressions of the same
mapping between object and data space in different bases. Figure
\ref{fig:monopole-block} summarizes the relationship between object and data
space in both bases.

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{../figures/monopole-block/monopole-block.pdf}
  \caption{We can compute the mapping between object space and data space using
    two different bases.}
     \label{fig:monopole-block}      
\end{figure}

\subsection{Dipole imaging in different bases \tb{50\%}}\label{sec:dipole}
Now we consider a microscope imaging a field of in-focus \textit{dipoles} by
recording the irradiance on a two-dimensional detector. A function that assigns
a real number to each point on a plane is not sufficient to specify a field of
dipoles because the dipoles can have different orientations. To represent the
object we need to extend object space to
$\mbb{U} = \mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ where $\mbb{S}^2$ is the
two-dimensional sphere (the usual sphere embedded in $\mbb{R}^3$).

In a delta function basis the object can be represented by a function
$f(\ro, \so)$ called the \textit{spatio-angular density}---the number of dipoles
at position $\ro{}$ per unit area oriented along $\so{}$ per unit solid angle.
If the imaging system is linear and shift invariant then the mapping between
object and data space can be expressed in a delta function basis as
  \begin{align}
g(\rd{}) = \int_{\mbb{S}^2}d\so{}\int_{\mbb{R}^2}d\ro{}\, h(\rd{} -\ro{}, \so{})f(\ro, \so), \label{eq:odpsf}
  \end{align}
  where $h(\rd - \ro, \so)$ is the \textit{orientation-dependent point spread
    function}.

  We can make our first change of basis by applying the Fourier-convolution
  theorem to Eq. \ref{eq:odpsf} which yields
\begin{align}
G(\bs{\nu}) = \int_{\mbb{S}^2}d\mh{s}\, H(\bs{\nu}, \mh{s})F(\bs{\nu}, \mh{s}) \label{eq:odotf},
\end{align}
where we define the \textit{irradiance spectrum} as
\begin{align}
  G(\bs{\nu}) &= \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
  \end{align}
  the \textit{orientation-dependent optical transfer function} as
  \begin{align}
  H(\bs{\nu}, \mh{s}) &= \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \mh{s})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
  \end{align}
  and the \textit{spatial density spectrum} as
  \begin{align}
  F(\bs{\nu}, \mh{s}) &= \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r}, \mh{s})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}). 
  \end{align}
  This basis is well suited for simulating objects that are angularly sparse and
  spatially dense.

  The spherical harmonics provide another set of convenient basis functions. We
  can change basis from spherical delta functions to spherical harmonics by
  making use of the generalized Plancharel theorem for spherical functions
\begin{align}
  \ints{2}d\mh{s}{}\, h(\mh{s})f(\mh{s}) = \lmsum H_\ell^m F_\ell^m, \label{eq:plan}
\end{align}
where
\begin{align}
  F_\ell^m \equiv \int_{\mbb{S}^2}d\mh{s}\, f(\mh{s})\bar{Y}_\ell^m(\mh{s}), 
\end{align}
is the spherical Fourier transform of $f(\mh{s})$, $\bar{z}$ denotes the complex
conjugate of $z$, and $Y_{\ell}^m(\mh{s})$ are the spherical harmonic functions
defined in Appendix \ref{sec:sph}. Eq. \ref{eq:plan} expresses the fact that
scalar products are invariant under a change of basis (see Eq. 3.78 of
\cite{barrett2004}). The left-hand side of Eq. \ref{eq:plan} is the scalar
product of $\mbb{L}_2(\mbb{S}^2)$ functions in a delta function basis and the
right hand side is the scalar product of $\mbb{L}_2(\mbb{S}^2)$ functions in a
spherical harmonic function basis. Applying Eq. \ref{eq:plan} to Eq.
\ref{eq:odpsf} yields
\begin{align}
  g(\rd) = \lmsum H_\ell^m(\rd - \ro)F_\ell^m(\ro) \label{eq:atf-form}
\end{align}
where we have defined the \textit{angular transfer function} as
\begin{align}
  H_\ell^m(\mb{r}) = \int_{\mbb{S}^2}d\mh{s}\, h(\mb{r}, \mh{s})\bar{Y}_{\ell}^m(\mh{s}),\label{eq:atf-prep} 
\end{align}
and the \textit{angular density spectrum} as
\begin{align}
  F_\ell^m(\mb{r}) = \int_{\mbb{S}^2}d\mh{s}\, f(\mb{r}, \mh{s})\bar{Y}_{\ell}^m(\mh{s}).
\end{align}
This basis is well suited for simulating objects that are spatially sparse and
angularly dense.

We can arrive at our final basis in two ways: by applying the generalized
Plancharel theorem for spherical functions to Eq. \ref{eq:odotf} or by applying
the Fourier convolution theorem to Eq. \ref{eq:atf-form}. We follow the
second path and find that
\begin{align}
G(\bs{\nu}) = \lmsum H_\ell^m(\bs{\nu})F_\ell^m(\bs{\nu}) \label{eq:saft},
\end{align}
where we have defined the \textit{spatio-angular transfer function} (SATF) as
  \begin{align}
  H_\ell^m(\bs{\nu}) &\equiv \int_{\mbb{S}^2}d\mh{s}\, H(\bs{\nu}, \mh{s})\bar{Y}_\ell^m(\mh{s}),
  \end{align}
  and the \textit{spatio-angular density spectrum} as
  \begin{align}
  F_\ell^m(\bs{\nu}) &\equiv \int_{\mbb{S}^2}d\mh{s}\, F(\bs{\nu}, \mh{s})\bar{Y}_\ell^m(\mh{s}).
  \end{align}
  This basis is well suited for simulating arbitrary samples because it exploits
  the sparsity of the imaging system---we will see this explicitly when we
  calculate a specific case of the spatio-angular transfer function.

  Figure \ref{fig:dipole-block} summarizes the relationship between the four
  bases that we can use to compute the image of a field of dipoles.
